### 2022-06-13 paper.tex
try to always use the same word for 100kb windows

Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution. Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Currently, this is simply done by tiling teh genome into large windows and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these windows. We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.

@@

Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
While a large 100~kb bin will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.

@@

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. \todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.

@@

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

\begin{figure*}
    \begin{center}
    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
    C\hspace*{.7\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
    D\hspace*{.7\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right). Data from \citet{luo2017single}.}
    \label{figure:score}
    \end{center}
\end{figure*}

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb bins, or by acquiring VMRs with \texttt{scbs scan} using default options. For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}. Intervals containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded. To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in. 
For dimensionality reduction, missing values were imputed with an iterative PCA as described above. To visualize data, UMAP was performed on the final 15 principal components.

>>>>

Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution. Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Currently, this is simply done by tiling the genome into large windows and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these windows. We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.

@@

Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
While a large 100~kb tile will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.

@@

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb windows or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. \todo{Is the following correct?} For both types of regions, the 100~kb windows and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.

@@

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic 100~kb windows is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

\begin{figure*}
    \begin{center}
    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
    C\hspace*{.7\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
    D\hspace*{.7\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right). Data from \citet{luo2017single}.}
    \label{figure:score}
    \end{center}
\end{figure*}

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb windows, or by acquiring VMRs with \texttt{scbs scan} using default options. For all regions (VMRs and 100~kb windows), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}. Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded. To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in. 
For dimensionality reduction, missing values were imputed with an iterative PCA as described above. To visualize data, UMAP was performed on the final 15 principal components.
